---
title: 'Order Statistics 1.3'
author: "Heather Ann Dye"
date: "11/8/2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Order Statistics 1.3

#####  Example 1: Definition of Order Statistics

Let $X_i$ be the observed value of the $i$th trial from a sequence of $n$ trials.
Let

* $Y_1 = min \lbrace X_1, X_2, \ldots X_n \rbrace$
* $Y_2 =$ second smallest $\lbrace X_1, X_2, \ldots X_n \rbrace$
* $Y_3 =$ third smallest $\lbrace X_1, X_2, \ldots X_n \rbrace$
* $Y_n = max \lbrace X_1, X_2, \ldots X_n \rbrace$

Now,
\[ Y_1 \leq Y_2 \leq Y_3 \ldots \leq Y_n. \]




The variables $Y_1 \leq Y_2 \leq Y_3 \leq Y_4 \leq Y_5$ are order statistics. In
this example,
each independent trial is a random variable $X_i$ with the distribution
\[ f(x_i) = \begin{cases} 2x_i \: & 0 \leq x \leq 1 \\ 0 \: & \text{ else } \end{cases} \]
Then 
\[ F(a_i)= \begin{cases} 0 \: & a_i \leq 0  \\ a_i^2 \: & 0 \leq a_i < 1 \\ 1 \: & 1 \leq a_i \end{cases} 
\]

Now, $P(X_i<1/2)=1/4$ so that $P(Y_4<1/2)= {5 \choose 4} (1/4)^4 (3/4)+ (1/4)^5$.
Note that if $Y_4$ is less than $1/2$ then $Y_1, Y_2, Y_3$ are also less than $1/2$.  
If $Y_5$ is less than $1/2$ then all five order statistics are less than $1/2$. 

Hence, we can write the following CDF for $Y_4$.
\[ G(a) = P(Y_4 < a) = \begin{cases} 0 & a < 0\\ {5 \choose 4} (a^2)^4 (1-a^2) + (a^2)^5 & 0 \leq a < 1\\  1 & 1 \leq a \end{cases} \]
This is the cumulative distribution function and the associated pdf is:
\[ G'(a)=g(a) = \begin{cases} 0 & a \leq 0  \\  {5 \choose 4} (4(a^2)^3 (2a) (1-a^2) -2a (a^2)^4) + 5(a^2)^4 (2a) & 0 \leq a < 1 \\  0 &  1 \leq a  \end{cases} \]

##### Example 2: Calculational Formula for the $r$th order statistic

Let $X_i$ be the $i$th sample of $n$ with the pdf $f(x)$ and the cdf $F(X)$. Then 
the CDF of $Y_r$ is
\[G_r (a) = Pr(Y_r < a) = \sum_{i=r} ^n {n \choose i} F(a)^i (1-F(a))^{n-i}. \]
with the appropriate domains. 
Then, we take the derivative and compute the pdf:
\[g_r(a)=\sum_{i=r} ^n {n \choose i}(i F(a)^{i-1}f(a) (1-F(a))^{n-i} + (n-i)F(a)^i (1-F(a))^{n-i-1} (-f(a))  ).\]


##### Example 3: Computing an order statistic. 

Let $X_i \sim N(0,1)$ be the $i$th sample of 6. Note that $f(x)$ is described by dnorm and $F(X)$ is described by pnorm  in R. 

We want to calculate the probability that $P(Y_5 < 1)$.

Note that 

$$P(Y_6 < a) = F(a)^6$$ 
and
$$P(Y_5 < a)= 6 F(a)^5(1-F(a)) + F(a)^6$$. 
We can easily compute this in R by writing a function. 



```{r}
myorderstat<-function(a){6*pnorm(a)^5 *(1- pnorm(a)) + pnorm(a)^6}
myorderstat(1)


```

##### Example 4: Setting up a Q-Q plot

A Q-Q plot is a quantile-quantile plot and compares order statistics with a given distribution. 
Note that $Y_r$ is the sample quantile of order $r/(n+1)$ and $\pi_r$ is the percentile with a Q-Q plot $(\pi_r, Y_r)$. In a quantile-quantile plot, the quantile of the statistic $Y_r$ is computed. Then the corresponding statistic in 
the comparison distribution is computed. If the two distributions are of the same
type, the ordered pairs will form a straight line since it is essentially a change of variables. 

##### Example 5:  Comparing Data with the normal distribution

In this example, data from a uniform distribution is compared to a normal distribution and to itself.
We first compare data from a uniform distribution to the normal distribution. 

```{r}
y=runif(200)
qqnorm(y, main = "Normal Q-Q Plot",
       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
       plot.it = TRUE, datax = FALSE)

qqline(y,  distribution = qnorm)
qqplot(y, y, xlab = "y vector", ylab = "y vector")


```

Take 2: a comparison with a normal distribution. 


```{r}
y=rnorm(200)
hist(y)
qqnorm(y, main = "Normal Q-Q Plot",
       xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",plot.it = TRUE, datax = FALSE)

qqline(y,  distribution = qnorm)






```





##### Example 6: Comparison of normal data with qqplot


We now compare normal data using the qqplot command in R. 
This is an alternative command, which will let us compare two 
data sets of our choice.


```{r}
x=rnorm(200)
x2=rnorm(200)
qqplot(x,x2)



```


###  Hand computation of Q-Q Plots

* Compute the size of the data set. 
* Rank order the data - the vector $x$
* For data in position $i$, compute $p=i/(n+1)$. The vector: $x2$
* Find $z$ such that $P(Z<z)=p$. The vector: $x3$
* Plot the sorted data versus the $x3$ vector

#### The idea
If $Y_r$ is $N( \mu, \sigma^2)$ then 
\[ z = \frac{y_r-\mu}{\sigma} \]
and 
\[ y_r = \mu + \sigma z. \]
If $Y_r$ represents the $k$th percentile then 
$Y_r = \mu + \sigma z_{1-r}$ where $P(Z\leq z_{1-r})=r$. 

##### Example 7: Normal data

We compute the q-q plot by hand in this example.

```{r}
x<-sort(rnorm(100, mean=5, sd=3))

x2<-seq(1/101, 100/101, by=1/101)
x3<-qnorm(x2)
plot(x,x3)
qqnorm(x)
```
##### Example 8: Uniform data - Not normally distributed. 

We compute the qq plot of uniform data.

```{r}
x=sort(runif(100))
x2=seq(1/101, 100/101, by=1/101)
x3=qnorm(x2)
plot(x,x3)
```

Compare your results to the output from qqnorm and qqline. 

```{r}
qqnorm(x); qqline(x)
```


##### Example 9: Uniform Data compared to a uniform distribution

We compare data from the uniform distribution to the uniform distribution.

```{r}
x=sort(runif(100))
x2=seq(1/101, 100/101, by=1/101)
x3=qunif(x2)
plot(x,x3)
```

Compare your calculation to the output of qqplot. 

```{r}
qqplot(x, x3); qqline(x, distribution = qunif)
```

## Basic Descriptive Statistics for samples 

Beyond percentiles and histograms.
We are work with a sample of size $n$. Assume that the population has size $N$. 
The theoretical distribution is unknown. Let $x_i$ denote the $i$th sample.
The term *population parameter* refers to the unknown parameters of the population. The *sample statistics* are calculated from the sample in an effort to estimate the parameters.

##### Example 10 - sample mean

The mean of a sample of size $n$ is denoted
$$\bar{x} = \sum_{i=1}^{n} \frac{x_i}{n}.$$
The notation $\mu$ is used for the mean of the population.

##### Example 11 - the sample variance and the sample standard deviation

The sample standard deviation is denoted as $s$ and the sample variance, $s^2$, is the square of the sample standard deviation. 
The sample variance is
$$s^2 = \sum_{i=1}^n \frac{(x_i - \bar{x})^2}{n-1}.$$
The variance is
$$ \sigma^2 = \sum_{i=1}^{N} \frac{(x_i - \mu)^2}{N}.$$

##### Example 12: The mode

The mode is the most frequently occurring value. 

##### Example 13: The empirical rule

The empirical rule is used to assess if a data set is bell shaped (or normal).
For a normal distribution:

* approximately 68% of the data falls in the interval $(\bar{x}-s, \bar{x}+s)$
* approximately 95% of the data falls in the interval $(\bar{x}-2s, \bar{x}+2s)$
* approximately 99.7% of the data falls in the interval $(\bar{x}-3s, \bar{x}+3s)$


#### Example 14: Simpson's paradox 

\[ \begin{matrix} 
 & Player A:AB & Hits & Average  & Player B:AB  & Hits & Average\\
\hline
Season 1 & 500 & 126 & 0.252&  300 & 75 & 0.250 \\
Season 2 & 300 & 90 & 0.300 &  500 & 145 & 0.290\\
\hline
Total & 800 & 216 & 0.270  & 800 & 220 & 0.275\end{matrix}\]


Notice that each season, Player A has a better average than Player B. 
However, over the two cumulative seasons, Player B has the better average. This is different from averaging the averages from each season. 


#### Example 16: R Demonstration of Descriptive Statistics

```{r}
set.seed(45)
mydata<-rbinom(40, 20, 0.5)
handmean<-sum(mydata)/length(mydata)
builtinmean<-mean(mydata)


handvariance<-sum((mydata-handmean)*(mydata - handmean))/(length(mydata)-1)
builtin<-sd(mydata)


emprule<-function(dvec, n){mean(dvec)+n*sd(dvec)*c(-1,1)}
emprule(mydata,2)

std1<-emprule(mydata,1)
instd1<-mydata[mydata>std1[1] & mydata<std1[2]]

length(instd1)/length(mydata)
mytable=table(mydata)
mytable
names(mytable)[mytable==max(mytable)]
```

