---
title: 'Unbiased Estimators - 1.5a'
author: "Heather Ann Dye"
date: "12/16/2022"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Unbiased Estimators

We can use both the method of moments or the MLE to estimate a parameter. Which is better? 




##### Example 1: Unbiased estimator

An **unbiased estimator** has the property that $E(\hat{\theta})=\theta$ where
$\hat{\theta}$ is an estimate of $\theta$. 

Let $X_i$ be a sample of size $n$ from a population with mean $\mu$. 
We estimate $\mu$:
\[ \hat{\mu}(x_1,x_2, \ldots x_n) = \frac{ \sum x_i}{n}. \]

This is an unbiased estimator since 
\[ E(\hat{\mu} ) = \frac{1}{n} E( \sum x_i) = \mu. \]

##### Example 2: Biased Estimator

Let $X_i$ be a sample of size $n$ drawn from a population with mean $0$ and variance $\sigma^2$. 
We compute
\[  E \left( \frac{ \sum (x_i - \bar{x})^2}{n} \right) \neq \sigma^2.\]

That is, this is a biased estimate of $\sigma^2$. 
The value 
\[  E \left( \frac{ \sum (x_i - \bar{x})^2}{n-1} \right) \]
is an unbiased estimate. 


##### Example 3: Simulation

In the following simulation, we can see that the population 
variance formula under estimates the variance. 


```{r}
mysample<-rnorm(3000)
mysample<-data.frame(matrix(data=mysample, nrow =30, ncol=100))
mymeans<-sapply(mysample, mean)
hist(mymeans)
mean(mymeans)


popvar<-function(myvec){sum((myvec-mean(myvec))^2)/length(myvec)}

popvar(c(2,3,5))


mypvar<-sapply(mysample, popvar)
hist(mypvar)
mean(mypvar)
mysamplesd<-sapply(mysample, sd)
mysamplevar<-(mysamplesd)^2

hist(mysamplevar)
mean(mysamplevar)
```


### Calculating the amount of error in the estimated parameter

To determine the amount of error in the estimate, $\hat{\theta}$, we calculate the estimated standard error, denoted as $s_{\hat{\theta}}$. 

##### Example 
For $i \in \lbrace 1, 2 \rbrace$, we denote the sample mean as $\bar{X_i}$ and the sample size as $n_i$. Let  $\mu_i$ and  $\sigma_i^2$ denote the population mean and variance respectively. 

Now, $\bar{X_1} + \bar{X_2}$ is an unbiased estimator since
$E(      \bar{X_1} + \bar{X_2}) = E(\bar{X_1}) + E(\bar{X_2}).$

The standard error of this estimate is:
\[ \sqrt{ Var(\bar{X_1} + \bar{X_2} )} = \sqrt{ Var(\bar{X_1}) + Var(\bar{X_2} )} =
\sqrt{ \frac{\sigma_1 ^2}{n_1} + \frac{\sigma_2 ^2 }{n_2} }. \]

The estimated standard error utilizes sample variance and is
\[ SE ( \bar{X_1} + \bar{X_2}) = \sqrt{ \frac{s_1 ^2}{n_1} + \frac{s_2 ^2 }{n_2} }. \] 



